Inference of Bifurcations with Differentiable Continuation

### 2021-02-22 docs/article.tex
update article

An expression for the field $\tangent(z)$ tangent to the set of curves would allow us to take derivatives and integrals along them. Fortunately the tangent field can be constructed by ensuring it is perpendicular to the gradient $\partial_z$ of each component of $\rates$ as illustrated by an example two component system in Figure \ref{fig:implicit-surfaces}. The tangent field $\tangent(z)$ can be constructed perpendicular to all gradient vectors using the properties of the determinant \cite{Goldman2005CurvatureSurfaces}
\begin{align}
    \tangent(z):=
    \label{eq:tangent-field}
    \left|\begin{matrix}
        \hat{z} \\
        \,\partial_{z}\rates\,
    \end{matrix}\right|
    \qquad\tangent : \Reals^{N+1}\rightarrow\Reals^{N+1}\\
    =\sum_{i=1}^{N+1}\hat{z}_{i}(-1)^{i+1} \left|\frac{\partial \rates}{\partial(z\setminus z_{i}) }\right|
\end{align}
where $\hat{z}$ is a collection of unit basis vectors in the $\Reals^{N+1}$ space and $\partial_{z}\rates$ is an $N\times(N+1)$ rectangular Jacobian matrix of partial derivatives and $z\setminus z_{i}$ denotes the $N$ dimensional vector $z$ with component $z_{i}$ removed. This construction ensures that the dot product of this field any gradient of a component of $\rates$
\begin{align}
    \tangent(z)\cdot\partial_z f_{\theta} =
    \left|\begin{matrix}
        \partial_z f_{\theta} \\
        \,\partial_{z}\rates\,
    \end{matrix}\right|
    \quad =0 \quad\forall f_{\theta}\in \rates
\end{align}
since the determinant of any matrix with two identical rows or columns is zero. Note that the tangent field $\tangent(z)$ is actually defined for all values of $z$ where adjacent field lines trace out other level sets where $\rates(z)\neq0$. Furthermore deformations with respect to $\theta$ are always orthogonal to the tangent \todo{numerically true. analytic proof?}
\begin{align}
    \tangent(z)\cdot\frac{d\tangent}{d\theta}=0
\end{align}

@@

\subsection{Minimal Models}
Fig

\subsection{Applied Models}

\subsubsection{Two-state model (mutual inhibition)}

\begin{equation}
    \dfrac{du_1}{dt} = \dfrac{a_1} { 1 + (p u_2)^2 } - \mu_1 u_1, \quad
    \dfrac{du_2}{dt} = \dfrac{a_2} { 1 + (k u_1)^2 } - \mu_2 u_2
\end{equation}
Without loss of generality, we can rescale $u_1$ by $a_1$ and $u_2$ by $a_2$:
\begin{equation*}
    \dfrac{du_1}{dt} = \dfrac{1} { 1 + (\hat{p} u_2)^2 } - \mu_1 u_1, \quad
    \dfrac{du_2}{dt} = \dfrac{1} { 1 + (\hat{k} u_1)^2 } - \mu_2 u_2
\end{equation*}
where $\hat{p} = p a_2$ and $\hat{k} = k a_1$. The equilibria for this system are defined by
\begin{equation}
    \mu_1u_1^*(1+\hat{p}u_2^*)^2 = 1, \quad \mu_2u_2^*(1+\hat{k}u_1^*)^2 = 1
\end{equation}
Obtaining closed form expressions for $u_1^*$ and $u_2^*$ requires solving a quintic polynomial. We won't do that here.

What is the determinant for this system?
\begin{equation*}
    \left|\dfrac{\partial F}{\partial u}\right| = 
    \begin{vmatrix} -\mu_1 & -\frac{2 p^2 u_2^*}{(1 + (p u_2^*)^2)^2} \\
        -\frac{2k^2u_1^*}{(1+(k u_1^*)^2)^2} & -\mu_2
    \end{vmatrix}
    = \mu_1\mu_2 - \frac{4p^2k^2u_1^*u_2^*}{(1+(p u_2^*)^2)^2(1+(k u_1^*)^2)^2}
\end{equation*}

\subsubsection{Two-state (generalised)}
\begin{equation}
    \dfrac{du_1}{dt} = \dfrac{a_1 + b_1(p u_2)^2} { 1 + (p u_2)^2 } - \mu_1 u_1, \quad
    \dfrac{du_2}{dt} = \dfrac{a_2 + b_2(k u_1)^2} { 1 + (k u_1)^2 } - \mu_2 u_2
\end{equation}

Applying the same rescaling as above, we can simplify as
\begin{equation}
    \dfrac{du_1}{dt} = \dfrac{1 + \rho_1(p u_2)^2} { 1 + (p u_2)^2 } - \mu_1 u_1, \quad
    \dfrac{du_2}{dt} = \dfrac{1 + \rho_2(k u_1)^2} { 1 + (k u_1)^2 } - \mu_2 u_2
\end{equation}
where $\rho_i = \frac{b_i}{a_i}$, with other parameters rescaled as above (but dropping hats).

@@

\section{Discussion}
\label{sec:discussion}

\subsection{Model Synthesis}

\subsection{Model Selection}

>>>>

An expression for the field $\tangent(z)$ tangent to the set of curves would allow us to take derivatives and integrals along them. Fortunately the tangent field can be constructed by ensuring it is perpendicular to the gradient $\partial_z$ of each component of $\rates$ as illustrated by an example two component system in Figure \ref{fig:implicit-surfaces}. The tangent field $\tangent(z)$ can be constructed perpendicular to all gradient vectors using the properties of the determinant \cite{Goldman2005CurvatureSurfaces}
\begin{align}
    \tangent(z):=
    \label{eq:tangent-field}
    \left|\begin{matrix}
        \hat{z} \\
        \,\partial_{z}\rates\,
    \end{matrix}\right|
    \qquad\tangent : \Reals^{N+1}\rightarrow\Reals^{N+1}\\
    =\sum_{i=1}^{N+1}\hat{z}_{i}(-1)^{i+1} \left|\frac{\partial \rates}{\partial(z\setminus z_{i}) }\right|
\end{align}
where $\hat{z}$ is a collection of unit basis vectors in the $\Reals^{N+1}$ space and $\partial_{z}\rates$ is an $N\times(N+1)$ rectangular Jacobian matrix of partial derivatives and $z\setminus z_{i}$ denotes the $N$ dimensional vector $z$ with component $z_{i}$ removed. This construction ensures that the dot product of this field any gradient of a component of $\rates$
\begin{align}
    \tangent(z)\cdot\partial_z f_{\theta} =
    \left|\begin{matrix}
        \partial_z f_{\theta} \\
        \,\partial_{z}\rates\,
    \end{matrix}\right|
    \quad =0 \quad\forall f_{\theta}\in \rates
\end{align}
since the determinant of any matrix with two identical rows or columns is zero. Note that the tangent field $\tangent(z)$ is actually defined for all values of $z$ where adjacent field lines trace out other level sets where $\rates(z)\neq0$. Furthermore deformations with respect to $\theta$ are always orthogonal to the tangent
% \todo{numerically true. analytic proof?}
\begin{align}
    \tangent(z)\cdot\frac{d\tangent}{d\theta}=0
\end{align}

@@

\subsection{Minimal Models}
Figures \ref{fig:saddle-node:results} and \ref{fig:pitchfork:results} show example optimisations of $(\theta_1,\theta_2)$ for the minimal saddle-node and pitchfork models respectively. The magnitude of the correction factor $\lambda=0$ when bifurcations are present and $\lambda\neq0$ otherwise.

\subsection{Applied Models}
In progress...
\begin{itemize}
    \item Genetic toggle switch
    \item FitzHugh-Nagumo
\end{itemize}
% \subsubsection{Genetic toggle switch}
% \subsubsection{FitzHugh-Nagumo}

% \begin{equation}
%     \dfrac{du_1}{dt} = \dfrac{a_1} { 1 + (p u_2)^2 } - \mu_1 u_1, \quad
%     \dfrac{du_2}{dt} = \dfrac{a_2} { 1 + (k u_1)^2 } - \mu_2 u_2
% \end{equation}
% Without loss of generality, we can rescale $u_1$ by $a_1$ and $u_2$ by $a_2$:
% \begin{equation*}
%     \dfrac{du_1}{dt} = \dfrac{1} { 1 + (\hat{p} u_2)^2 } - \mu_1 u_1, \quad
%     \dfrac{du_2}{dt} = \dfrac{1} { 1 + (\hat{k} u_1)^2 } - \mu_2 u_2
% \end{equation*}
% where $\hat{p} = p a_2$ and $\hat{k} = k a_1$. The equilibria for this system are defined by
% \begin{equation}
%     \mu_1u_1^*(1+\hat{p}u_2^*)^2 = 1, \quad \mu_2u_2^*(1+\hat{k}u_1^*)^2 = 1
% \end{equation}
% Obtaining closed form expressions for $u_1^*$ and $u_2^*$ requires solving a quintic polynomial. We won't do that here.

% What is the determinant for this system?
% \begin{equation*}
%     \left|\dfrac{\partial F}{\partial u}\right| = 
%     \begin{vmatrix} -\mu_1 & -\frac{2 p^2 u_2^*}{(1 + (p u_2^*)^2)^2} \\
%         -\frac{2k^2u_1^*}{(1+(k u_1^*)^2)^2} & -\mu_2
%     \end{vmatrix}
%     = \mu_1\mu_2 - \frac{4p^2k^2u_1^*u_2^*}{(1+(p u_2^*)^2)^2(1+(k u_1^*)^2)^2}
% \end{equation*}

% \subsubsection{Two-state (generalised)}
% \begin{equation}
%     \dfrac{du_1}{dt} = \dfrac{a_1 + b_1(p u_2)^2} { 1 + (p u_2)^2 } - \mu_1 u_1, \quad
%     \dfrac{du_2}{dt} = \dfrac{a_2 + b_2(k u_1)^2} { 1 + (k u_1)^2 } - \mu_2 u_2
% \end{equation}

% Applying the same rescaling as above, we can simplify as
% \begin{equation}
%     \dfrac{du_1}{dt} = \dfrac{1 + \rho_1(p u_2)^2} { 1 + (p u_2)^2 } - \mu_1 u_1, \quad
%     \dfrac{du_2}{dt} = \dfrac{1 + \rho_2(k u_1)^2} { 1 + (k u_1)^2 } - \mu_2 u_2
% \end{equation}
% where $\rho_i = \frac{b_i}{a_i}$, with other parameters rescaled as above (but dropping hats).

@@

% \section{Discussion}
% \label{sec:discussion}

% \subsection{Model Synthesis}
% \subsection{Model Selection}
